---
title: "Exploring Hindu Muslim Violence in India"
author: "Ojaswi Malik"
output: github_document
---

## Introduction

### Goal: This report aims to explore the history of violent events between the 2 controversial religions of Hinduism and Islam in the context of post-independant India. Keeping in mind the recent riots in the capital New Delhi, it is essential to trace back and analyse this relationship in the latter half of the 20th century. This report relies on the dataset created by Varshney and Wilkinson, who collected all incidents on Hindu-Muslim Violence in India between 1950-1995 reported in the national newspaper Times of India. This report will 




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#loading necessary libraries
library(tidyverse)
library(dplyr)
library(ggplot2)
library(here)


# set default theme
theme_set(theme_minimal())
```

## Importing the data files

```{r import data}
#Importing the first dataset
hindu_muslim_data <- here("data", "hindu_muslim_violence.csv") %>%
  read_csv()

#Importing the second dataset and parsing column types
yearly_pop_pm_data <- here("data", "yearly_pop_pm.csv") %>%
  read_csv( col_types = 
              cols(
                hindu_prop = col_double(),
                muslim_prop = col_double()
          
              ))
```

## Data Wrangling

``` {r wrangling}
#Joining the 2 datasets using left join
hindu_muslim_complete_data <- left_join(
                                x = hindu_muslim_data, 
                                y = yearly_pop_pm_data
                                      )

```

## Exploring Data

### **Analysis 1**: First, I want to see how many hindu-muslim violence incidents were reported each year from 1950-1995

```{r count yearly incidents}

hindu_muslim_complete_data %>%
  #grouping by year to count incidents per year
  group_by(YEAR) %>%
  summarise( 
    count = n()
    ) %>%
  #plotting the line chart
  ggplot(mapping = aes(x = YEAR, y = count)) +
  geom_line() +
  #labelling the graph
  labs(
    title = "Number of Hindu-Muslim violence incidents from 1950-1995" ,
    x = "Year",
    y = "Number of incidents",
    caption = "Source: Varshney, Wilkinson dataset"
  )

```

### **Observation 1**: From the above graph, we can pbserve that yearly violent events are un-related and can be extremely high in one year, and low in the next. However, there was a peak in violent events in 1985 and 1990, which can be explained by the communal riots in that period. Moreover, in noticing a general trend, there has been a rise in violent events over the years. 

### **Analysis 2**: Now, I want to see the number of violent events aggregated by month. This is to observe if any month would have an unusual trend due to seasonal or festivities causes.

```{r monthy count incidents}

hindu_muslim_complete_data %>%
  ggplot( mapping = aes(x = MONTH_BY_N)) +
  geom_bar() +
  scale_x_discrete(limits = month.abb ) +
   labs(
    title = "Monthly number of Hindu-Muslim violence incidents from 1950-1995" ,
    x = "Month",
    y = "Number of incidents",
    caption = "Source: Varshney, Wilkinson dataset"
  )


```

### **Observation 2**: March has the highest number of violence incidents in the entire year. However, it is closely followed by October and September.

### **Analysis 3**: In this analysis, I want to observe the state-wise number of Hindu-Muslim incidents of violents

```{r state-wise}

hindu_muslim_complete_data %>%
  #filtering out NA values
  filter( !is.na(STATECODE)) %>%
  group_by(STATECODE) %>%
  summarise(
    count = n()
  ) %>%
  ggplot(mapping = aes(y = count, x = fct_reorder(STATECODE, count))) +
  geom_col() + 
  coord_flip() +
  labs(
    title = "State-wise Hindu-Muslim violence incidents from 1950-1995" ,
    y = "Number of Incidencts",
    x = "State",
    caption = "Source: Varshney, Wilkinson dataset"
  )

```

### **Observation 3**: The state of GU ie Gujarat has the highest number of incidents of Hindu-Muslim violence. MA (Maharashtra) and UP (Uttar Pradesh) also have extremely high number of incidents

### **Note regarding Analysis 3**: In constructing this analysis, I noticed the fallacies in coding the data by the authors. Initially, I had tried to use STATE to get full state-names, but some states were written with different spellings or had another tiny differences which lead to similiar states being printed multiple times. However, the use of STATECODE resolved this error.

### **Analysis 4**: In the top 3 states with the highest number of incidents as observed in Analysis 3, I want to see the distribution of number of killed and the number of arrests

```{r top-3 distribution }

hindu_muslim_complete_data %>%
  #only looking at the top 3 states of GU, MA and UP
  filter( STATECODE == c('GU', 'MA', 'UP')) %>%
  group_by(STATECODE) %>%
  

```

### **Observation 4**:

## Observations

## Conclusion

