---
title: "Exploring Hindu Muslim Violence in India"
author: "Ojaswi Malik"
output: github_document

---

## Introduction

### Goal: 
This report aims to explore the history of violent events between the 2 controversial religions of Hinduism and Islam in the context of post-independant India. Keeping in mind the recent riots in the capital New Delhi, it is essential to trace back and analyse this relationship in the latter half of the 20th century. This report relies on the dataset created by Varshney and Wilkinson, who collected all incidents on Hindu-Muslim Violence in India between 1950-1995 reported in the national newspaper Times of India. 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#loading necessary libraries
library(tidyverse)
library(dplyr)
library(ggplot2)
library(knitr)
library(here)

# set default theme
theme_set(theme_minimal())
```

### Importing the data files

```{r import data, message=FALSE, warning=FALSE}
#Importing the first dataset
hindu_muslim_data <- here("data", "hindu_muslim_violence.csv") %>%
  read_csv()

#Importing the second dataset and parsing column types
yearly_pop_pm_data <- here("data", "yearly_pop_pm.csv") %>%
  read_csv( col_types = 
              cols(
                hindu_prop = col_double(),
                muslim_prop = col_double()
              ))
```

### Data Wrangling

``` {r wrangling, message=FALSE, warning=FALSE}
#Joining the 2 datasets using left join
hindu_muslim_complete_data <- left_join(
                                x = hindu_muslim_data, 
                                y = yearly_pop_pm_data
                                      )
```

## Exploring Data

### **Analysis 1**: 
First, I want to see how many hindu-muslim violence incidents were reported each year from 1950-1995

```{r count yearly incidents}

hindu_muslim_complete_data %>%
  #grouping by year to count incidents per year
  group_by(YEAR) %>%
  summarise( 
    count = n()
    ) %>%
  #plotting the line chart
  ggplot(mapping = aes(x = YEAR, y = count)) +
  geom_line() +
  #labelling the graph
  labs(
    title = "Number of Hindu-Muslim violence incidents from 1950-1995" ,
    x = "Year",
    y = "Number of incidents",
    caption = "Source: Varshney, Wilkinson dataset"
  ) 

```

### **Observation 1**: 
From the above graph, we can pbserve that yearly violent events are un-related and can be extremely high in one year, and low in the next. However, there was a peak in violent events in 1985 and 1990, which can be explained by the communal riots in that period. Moreover, in noticing a general trend, there has been a rise in violent events over the years. 

### **Analysis 2**: 
Next I want to observe the changes in Hindus and Muslims proportion of population in the time period 1950-1995.

```{r proportion population, message=FALSE, warning=FALSE}
#Saving this graph to reuse in next analysis
prop_change <- hindu_muslim_complete_data %>%
  #grouping by year
  group_by(YEAR) %>%
  #plotting the graph
  ggplot(mapping = aes(x = YEAR)) + 
  geom_line(mapping = aes(y = hindu_prop, color = "Proportion of Hindu population")) +
  geom_line(mapping = aes(y = muslim_prop, color = "Proportion of Muslim population")) +
  #making y-axis in %
  scale_y_continuous(labels = scales::percent) +
  #labelling graph
  labs(
    title = "Proportion of Hindu and Muslim population in India  from 1950-1995" ,
    x = "Year",
    y = "Percentage",
    caption = "Source: Socio-Economic and Caste Census, 2011",
    color = NULL
  ) +
  #moving the legend to the bottom
  theme(legend.position = "bottom")
  
prop_change
```

### **Observation 2**: 
Although the changes have been small, we can observe a gradual decrease in proportion of Hindus and an increase in the proportion of Muslims as recorded in the census data between 1950-1995. 

### **Analysis 3**: 
Now, I want to see the number of violent events aggregated by month. This is to observe if any month would have an unusual trend due to seasonal or festivities causes.

```{r monthy count incidents}

hindu_muslim_complete_data %>%
  ggplot( mapping = aes(x = MONTH_BY_N)) +
  geom_bar() +
  #labelling and setiing limits to months on x-axis
  scale_x_discrete(limits = month.abb ) +
  #labelling graph
   labs(
    title = "Monthly number of Hindu-Muslim violence incidents from 1950-1995" ,
    x = "Month",
    y = "Number of incidents",
    caption = "Source: Varshney, Wilkinson dataset"
  )


```

### **Observation 3**: 
March has the highest number of violence incidents in the entire year. However, it is closely followed by October and September.

### **Analysis 4**: 
In this analysis, I want to observe the state-wise number of Hindu-Muslim incidents of violents

```{r state-wise}

hindu_muslim_complete_data %>%
  #filtering out NA values
  filter( !is.na(STATECODE)) %>%
  group_by(STATECODE) %>%
  summarise(
    count = n()
  ) %>%
  ggplot(mapping = aes(y = count, x = fct_reorder(STATECODE, count))) +
  geom_col() + 
  #flipping coordinates to see each state clearly
  coord_flip() +
  #labelling graph
  labs(
    title = "State-wise Hindu-Muslim violence incidents from 1950-1995" ,
    y = "Number of Incidents",
    x = "State",
    caption = "Source: Varshney, Wilkinson dataset"
  )

```

### **Observation 4**: 
The state of GU ie Gujarat has the highest number of incidents of Hindu-Muslim violence. MA (Maharashtra) and UP (Uttar Pradesh) also have extremely high number of incidents

###**Note regarding Analysis 4**: 
In constructing this analysis, I noticed the fallacies in coding the data by the authors. Initially, I had tried to use STATE to get full state-names, but some states were written with different spellings or had another tiny differences which lead to similiar states being printed multiple times. However, the use of STATECODE resolved this error.

### **Analysis 5**: 
In this analysis, I want to focus on the top 3 states observed above i.e. Gujarat, Maharashtra and Uttar Pradesh. Specifically I want to see the distribution of the number of people killed in an incident for the number of incidents.

```{r top-3 distribution, message=FALSE, warning=FALSE }

#To modify facet label text later
state.labs <- c("Gujarat", "Maharashtra", "Uttar Pradesh")
names(state.labs) <- c("GU", "MA", "UP")

hindu_muslim_complete_data %>%
  #filtering only the top 3 states
  filter(STATECODE == c("GU","MA","UP")) %>%
  ggplot(mapping = aes(x = KILLED)) +
  geom_histogram() +
  #Modifying the label text of the facets by replacing state code with state name
  facet_wrap( ~STATECODE, scales = "free", ncol = 1, labeller = labeller(STATECODE = state.labs)) +
  #labelling graph
  labs(
    title = "Hindu-Muslim violence incidents in top-3 states from 1950-1995" ,
    y = "Number of Incidents",
    x = "Number of people killed",
    caption = "Source: Varshney, Wilkinson dataset"
  )
  

```

### **Observation 5**: 
Even though Gujarat has the highest number of violence incidents, it does not have the highest number of killed people. Maharashtra has the largest variance due to having the maximum in the number of killed exceeding 400 people. This is followed by Uttar Pradesh and then Gujarat.

### **Analysis 6**: 
This analysis aims to see the effect of duration of a violent incident on the number of deaths

```{r duration effect, message=FALSE, warning=FALSE}
 hindu_muslim_complete_data %>%
  #grouping by duration of incident
  group_by(DURATION_I) %>%
  #Calculating the mean number of people killed and arrested for an incident with mean duration in days
  summarise(
    duration = mean(DURATION_I, na.rm = TRUE),
    killed = mean(KILLED, na.rm = TRUE),
    arrested = mean(ARRESTS, na.rm = TRUE)
  ) %>% 
  #removing the values which were calculated as NaN
  filter( !is.nan(killed)) %>%
  filter( !is.nan(arrested)) %>%
  #layering both line graphs to depict killed and arrested 
  ggplot(mapping = aes(x = duration)) + 
  geom_line(mapping = aes(y = killed, color = "Killed")) +
  geom_line(mapping = aes(y = arrested, color = "Arrested")) +
  #labelling the graph
  labs(
    title = "Hindu-Muslim violence incidents from 1950-1995" ,
    x = "Duration of Incident (days)",
    y = "Number of people killed/arrested",
    caption = "Source: Varshney, Wilkinson dataset",
    color = NULL
  ) +
  #moving the legend to the bottom
  theme(legend.position = "bottom")

```

### **Observation 6**: 
There does not seem to be any direct relationship between the duration of the incident and the corresponding number of deaths or arrests. However, a peak in both arrests and killed people is observed when the duration of the incident is between 7-10 days.

### **Analysis 7**: 
This analysis aims to find if reports of Dalit vs Muslims violence during the riot led to higher arrests

#### **Note**: 
Dalits (meaning "broken/scattered" in Sanskrit), is a term used for those aboriginal ethnic groups who have been subjected to untouchability.

```{r Dalits and arrests, message=FALSE, warning=FALSE}
 
hindu_muslim_complete_data %>%
  ggplot(mapping = aes(x = DALIT_MUSL, y = ARRESTS)) + 
  #constructing a boxplot
  geom_boxplot() +
  #labelling the graph
  labs(
    title = "Hindu-Muslim violence incidents from 1950-1995" ,
    x = "Reports of Dalit vs Muslim violence during the riot",
    y = "Number of people arrested",
    caption = "Source: Varshney, Wilkinson dataset"
  )
  
```

### **Observation 7**: 
The median number of arrests are clearly higher when Dalit vs Muslim violence is reported during the riot. However, for when Dalit vs Muslim violence is not reported, there are a larger number of outliers and larger variance. This can also be because the data set has more "No" reported cases than "Yes".

### **Analysis 8**: 
In this analysis, I want to narrow down the the first analysis we did, and look at it from the lens of the political leadership in place. I first want to observe the Prime Ministers in power between the years 1950-1995.

#### **Analysis 8.1**: 
This analysis constructs a facet wrap depicting the incidents of Hindu-Muslim violence occurring when different prime ministers were in power between 1950-1995.

```{r prime-minister1, message=FALSE, warning=FALSE}

number_incidents <- hindu_muslim_complete_data %>%
  #filtering out the missing prime minister data
  filter(!is.na(prime_minister)) %>%
  #grouping by year to count incidents per year
  group_by(YEAR, prime_minister) %>%
  summarise( 
    count = n()
    ) %>%
  #plotting the line chart
  ggplot(mapping = aes(x = YEAR, y = count)) +
  geom_line() +
  #labelling the graph
  labs(
    title = "Prime minister wise Number of Hindu-Muslim violence incidents from 1950-1995" ,
    x = "Year",
    y = "Number of incidents",
    caption = "Source: Varshney, Wilkinson dataset"
  ) 

number_incidents + facet_wrap(~prime_minister)

```

#### **Observation 8.1**: 
From this facet wrap, we can observe a noticeable graph only for 4 prime minister.s. This is because of the relatively short terms the other prime ministers had (ranging between 13 days and less than year)

#### **Analysis 8.2**: 
Hence, in this next analysis I want to filter only those 4 prime ministers and narrow down this data

```{r prime-minister2}
four_pms <- c("Indira Gandhi", "Jawahar Lal Nehru", "P.V. Narasimha Rao", "Rajiv Gandhi")

hindu_muslim_complete_data %>%
  #Observing only the 4 prime ministers with longer tenures
  filter( prime_minister == four_pms) %>%
  #grouping by year to count incidents per year
  group_by(YEAR, prime_minister) %>%
  summarise( 
    count = n()
    ) %>%
  #plotting the line chart
  ggplot(mapping = aes(x = YEAR, y = count)) +
  geom_line() +
  #labelling the graph
  labs(
    title = "Prime Minister wise number of Hindu-Muslim violence incidents from 1950-1995" ,
    x = "Year",
    y = "Number of incidents",
    caption = "Source: Varshney, Wilkinson dataset"
  ) +
  facet_wrap(~ prime_minister)

```

#### **Observation 8.2**: 
From the above facetted graph, we can observe that under the power of Rajiv Gandhi (among the 4 prime ministers we narrowed down to), there were the largest reported incidents of Hindu-Muslim violence.

#### **Analysis 8.3**: 
Finally, I want to find out the political parties these prime ministers were affiliated to while in power.

```{r political party, message=FALSE, warning=FALSE}

#Filtering only the 4 prime ministers we observed in our inital graph
pm_party <- hindu_muslim_complete_data %>%
  distinct(prime_minister, party) %>%
  filter(prime_minister == four_pms) %>%
  select(prime_minister, party)

# Column Names for the kable()
column_names <- c('Prime Minister', 'Party') 

#constructing the table
knitr::kable(
  pm_party,
  format = "html",
  caption = "Prime Ministers and their party affiliations",
  col.names = column_names,
)


```

#### **Observation 8.3**: 
From this table, we can notice that all 4 prime ministers belonged to the Indian National Congress.

## Concluding Observations:

This report has explored the dataset constructed by Varshney and Wilkinson along multiple variables to generate a deep insight into the reported Hindu-Muslim incidents of violence in the setting of post-independant India between 1950-1995. 

- **Overview**: 

  - While there was not a general trend in the number of violent incidents, there were certainly 2 mentionable peaks in late 1980 and early 1990. This can be explained by the infamous 1990 communal riots situated in Hyderabad, India. Moreover, according to the census data which is collected once every 10 years, the proportion of Hindus has decreased and proportion of Muslims has decreased between 1950-1995. However, this change is extremely small.
  
  - Month-wise, the month of March had the highest number of reported incidents in the given time period. This could be due to the celebration of the big Hindu festival of Holi in that month. However, this is only a speculation, and there could be other reasons.

- **State-Wise Overview**:

  - State-wise, the western state of Gujarat had the highest number of reported incidents by a large margin. Interestingly, the same state is also infamous for the Hindu-Muslim riots in 2002, which is not even recorded in this dataset. Gujarat, is followed by Maharashtra and Uttar Pradesh which are close in number.
  
  - Looking at these 3 states, Uttar Pradesh had the largest number of people killed as a result of the incidents of violence. Moreover, it also has the largest variance.
  

- **Arrested and Killed**

  - Before observing the effect of the duration of a given incident, one would expect a longer lasting incident to have larger number of deaths or arrests. However, on plotting the graph, there was no noticeable trend between these 3 variables, except for a peak for number of arrests due to an outlier case.
  
- **Dalit vs Muslims**

  - Since decades, dalits have bared the wrath of the Hindu hierarchical society. Literally translating to untouchable, they were considered to be at the bottom of the 5 "varna" system in Hinduism. Hence, I wanted to observe if involvement of Dalits in the violent acts against Muslims led to a change in number of arrests. Although the median number of arrests were higher if Dalit vs Muslim violence was reported, however there was a larger variance in incidents they were not reported.
  
- **Hindu Muslim Violence under different Prime Ministers**

  - I narrowed by observations of prime ministers to 4 prime ministers who had relatively longer tenures than their counterparts to get noticeable graphs when plotted to see number of incidents. These 4 prime minisiters were Indira Gandhi, Jawahar Lal Nehru, P.V. Narasimha Rao and .Rajiv Gandhi. 
  
  - Under Rajiv Gandhi's tenure, the highest number of Hindu-Muslim violent incidents were reported.
  
  - Interestingly, all 4 prime ministers belonged to the same party i.e. the Indian National Congress. However, this is not a big surprise since the Indian National Congress had majority for decades after independance.
  
  
- **Conclusion**: Although this topic is extremely sensitive, especially after the introduction of the discriminatory NRC and CAA acts recently in India, it is essential to approach this topic with an open mind and consider the multiple variables in the past that have led to its current situation.

## Limitation

I was unable to analyse the variable 'LOCAL_PREC' due to the way it was coded by the authors in the data. Many similiar observations were worded differently or spelled differently, leading to repition. This is exemplified in counting number of (LOCAL_PREC) where there are 129 different reasons leading to an incident, but there are many incidents with the same reason. This meant I was not able to analyse the underlying reasons behind different riots. I am looking forward to learning ways I can manipulate around this fallacy.

```{r limitation demo, echo = FALSE}
hindu_muslim_complete_data %>%
  distinct(LOCAL_PREC)
```
